LaMachine: A meta-distribution for NLP software

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] further updates

LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS. Second, there is flexibility with
regard to the form, where we support \emph{containerisation} through Docker\footnote{\url{https://www.docker.com}},
\emph{virtualisation} through Vagrant\footnote{\url{https://vagrant.org}} and VirtualBox\footnote{\url{https://www.virtualbox.org}},
direct remote provisioning through Ansible (for production servers), or an installation that is either global to the
machine or local in a custom directory for a specific user. Pre-built docker containers and virtual machine images with a
limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively.

@@

LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
repositories\footnote{i.e. the repositories where tool providers deposit their software} to the CodeMeta
standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
encouraging software developers to provide their own. This in turn enables other tools to do proper service discovery
and provenance logging.

Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
access and overview of all installed tools, including web services and web applications. It also comes with a
Jupyter Lab\footnote{\url{https://jupyter.org/}} environment which provides a web-based Integrated Development
Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
text and code and have gained great popularity in data science community nowadays.

@@

LaMachine exists since May 2015 and has been used extensively ever since by numerous users. It was initiallly conceived
as the primary means of distribution of the software stack developed at CLST, Radboud University Nijmegen, it includes a
lot of our software.  A full list of included software goes beyond the scope of this overview; we will merely mention
some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic
Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction
pipeline) and CLAM. We already include some relevant software by the University of Groningen (such as Alpino, a
dependency parser) and VU Amsterdam as well. Moreover, LaMachine incorporates a large number of renowned tools by
external international parties, offering most notably a mature Python environment with renowned scientific modules such
as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others, but also R, Java and
tools such as Stanford CoreNLP.

>>>>

LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS (although with more limitations). Second, there is
flexibility with regard to the form, where we support \emph{containerisation} through
Docker\footnote{\url{https://www.docker.com}}, \emph{virtualisation} through Vagrant\footnote{\url{https://vagrant.org}}
and VirtualBox\footnote{\url{https://www.virtualbox.org}}, direct remote provisioning through Ansible (for production
servers), or an installation that is either global to the machine or local in a custom directory for a specific user
(using \texttt{virtualenv}). Pre-built docker containers and virtual machine images with a limited selection of participating software are regularly
uploaded to the Docker Hub and Vagrant Cloud, respectively. The different flavours all offer a different degree of
seperation from the host OS, where Virtual Machines are completely virtualised, Docker Containers still share the kernel
with the host OS, and the machine-specific installation flavour actually compiles against the machine's distribution itself and thus
offers the least amount of overhead.

@@

LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
repositories, i.e. the repositories where tool providers deposit their software, to a common standard called CodeMeta
\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
encouraging software developers to provide their codemeta metadata inside their source code repositories and using
that directly. This in turn enables other tools to do proper service discovery and provenance logging.

Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
access and overview of all installed tools, including web services and web applications. It also comes with a
Jupyter Lab\footnote{\url{https://jupyter.org/}} environment which provides a web-based Integrated Development
Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
text, code and data output and have gained great popularity in data science community nowadays.

@@

LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in 2018 version 2 was
released which was a significant rewrite. LaMachine was initially conceived as the primary means of distribution of the
software stack developed at CLST, Radboud University Nijmegen. It therefore includes a lot of our software.  A full list
of included software goes beyond the scope of this overview; we will merely mention some CLARIN-NL/CLARIAH-funded tools:
ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic Annotation, with assorted tools), FLAT
(a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction pipeline) and CLAM. However, this
project is not limited to one research group and is open to participation by other software providers, especially those
also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by the University of
Groningen (such as Alpino, a dependency parser) and VU Amsterdam. Moreover, LaMachine incorporates a large number of
renowned tools by external international parties, offering most notably a mature Python environment with renowned
scientific modules such as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others,
but also R, Java and tools such as Stanford CoreNLP.
